---
title: "Statistical Distributions in R"
format: 
  html:
    code-fold: true
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    geometry:
      - top=15mm
      - bottom=20mm
      - heightrounded
---

While writing the [Randomness Concept][concept-randomness] for the Exercism learning syllabus, there was some uncertainty about what all the various R functions do.
Here, we demonstrate several them in action.

The `q*()` functions, to generate the quantiles, are not (yet) included.

First, we need the Tidyverse libraries and some default settings:

```{r}
#| title: imports
#| warning: false

library(tidyverse)

set_theme(theme_gray(base_size = 16))
```

## Normal/Gaussian Distribution

This is the classic "bell-shaped curve".

### The Normal Probability Density Function

The `dnorm()` function calculates the PDF at a point: the height of the curve at that point.

Different values for standard deviation are shown.
Higher values lead to wider, flatter curves.

All means default to zero.

```{r}
#| title: normal PDF

tibble(
  "x" = seq(-4, 4, length.out = 1000),
  "SD = 0.5" = dnorm(x, sd = 0.5),
  "SD = 1" = dnorm(x, sd = 1),
  "SD = 2" = dnorm(x, sd = 2)
)  |>
  pivot_longer(
    cols = starts_with("SD = "),
    names_to = "SD",
    values_to = "value"
  ) |>
  ggplot(aes(x = x, y = value, color = SD)) +
    geom_line() +
    facet_wrap(~SD, ncol = 1) +
    theme(legend.position = "none")
```

### The Normal Cumulative Distribution Function

The `pnorm()` function calculates the CDF at a point: the integral of the curve to the left of that point.

```{r}
#| title: normal CDF

tibble(
  "x" = seq(-4, 4, length.out = 1000),
  "SD = 0.5" = pnorm(x, sd = 0.5),
  "SD = 1" = pnorm(x, sd = 1),
  "SD = 2" = pnorm(x, sd = 2)
)  |>
  pivot_longer(
    cols = starts_with("SD = "),
    names_to = "SD",
    values_to = "value"
  ) |>
  ggplot(aes(x = x, y = value, color = SD)) +
    geom_line() +
    facet_wrap(~SD, ncol = 1) +
    theme(legend.position = "none")
```

### Generating random values with `rnorm()`

The `rnorm()` function generates random values which are normally distributed.

Here, we:

- get 10,000 values
- use a frequency polygon to bin them
- plot counts of the number in each bin on the `y` axis.

```{r}
#| title: random normal
#| warning: false

n <- 10000
tibble(
  "SD = 0.5" = rnorm(n, sd = 0.5),
  "SD = 1" = rnorm(n, sd = 1),
  "SD = 2" = rnorm(n, sd = 2)
)  |>
  pivot_longer(
    cols = starts_with("SD = "),
    names_to = "SD",
    values_to = "value"
  ) |>
  ggplot(aes(value, color = SD)) + 
    geom_freqpoly(binwidth = 0.05) +
    xlim(-4, 4) +
    facet_wrap(~SD, ncol = 1) +
    theme(legend.position = "none")
```

## Binomial Distribution

Here we use the example of 20 coin flips, counting the number that come up heads.

Three probabilities are used:

- `p = 0.4` for a coin biased to tails.
- `p = 0.5` for a fair coin.
- `p = 0.6` for a coin biased to heads.

### The Binomial Probability Density Function

```{r}
#| title: Binomial PDF

tibble(
  "x" = 0:20,
  "p = 0.4" = dbinom(x, size = 20, prob = 0.4),
  "p = 0.5" = dbinom(x, size = 20, prob = 0.5),
  "p = 0.6" = dbinom(x, size = 20, prob = 0.6)
)  |>
  pivot_longer(
    cols = starts_with("p = "),
    names_to = "p",
    values_to = "value"
  ) |>
  ggplot(aes(x = x, y = value, color = p)) +
    geom_col() +
    facet_wrap(~p, ncol = 1) +
    theme(legend.position = "none")
```

### The Binomial Cumulative Distribution Function

```{r}
#| title: Binomial CDF

tibble(
  "x" = 0:20,
  "p = 0.4" = pbinom(x, size = 20, prob = 0.4),#| warning: false

  "p = 0.5" = pbinom(x, size = 20, prob = 0.5),
  "p = 0.6" = pbinom(x, size = 20, prob = 0.6)
)  |>
  pivot_longer(
    cols = starts_with("p = "),
    names_to = "p",
    values_to = "value"
  ) |>
  ggplot(aes(x = x, y = value, color = p)) +
  geom_col() +
  facet_wrap(~p, ncol = 1) +
  theme(legend.position = "none")
```

### Generating random values with `rbinom()`

```{r}
#| title: random binomial

n <- 10000
tibble(
  "p = 0.4" = rbinom(n, size = 20, prob = 0.4),
  "p = 0.5" = rbinom(n, size = 20, prob = 0.5),
  "p = 0.6" = rbinom(n, size = 20, prob = 0.6)
)  |>
  pivot_longer(
    cols = starts_with("p = "),
    names_to = "p",
    values_to = "value"
  ) |>
  ggplot(aes(value, color = p)) + 
  geom_bar() +
  facet_wrap(~p, ncol = 1) +
  theme(legend.position = "none")
```

## Poisson Distribution

Here we use the example of how many meteorites at least 1m in diameter strike the Earth each year.

The `lambda` values are different assumptions for the _average_ number.
The [Catalina Sky Survey][web-catalina] discuss this and estimate such strikes happen "every few months", so `lambda = 4` is a reasonable guess.

### The Poisson Probability Density Function

```{r}
#| title: Poisson PDF

tibble(
  "x" = 0:15,
  "lambda = 1" = dpois(x, lambda = 1),
  "lambda = 4" = dpois(x, lambda = 4),
  "lambda = 8" = dpois(x, lambda = 8)
)  |>
  pivot_longer(
    cols = starts_with("lambda = "),
    names_to = "lambda",
    values_to = "value"
  ) |>
  ggplot(aes(x = x, y = value, color = lambda)) +
    geom_col() +
    facet_wrap(~lambda, ncol = 1) +
    theme(legend.position = "none")
```

### The Poisson Cumulative Distribution Function

```{r}
#| title: Poisson CDF

tibble(
  "x" = 0:15,
  "lambda = 1" = ppois(x, lambda = 1),
  "lambda = 4" = ppois(x, lambda = 4),
  "lambda = 8" = ppois(x, lambda = 8)
)  |>
  pivot_longer(
    cols = starts_with("lambda = "),
    names_to = "lambda",
    values_to = "value"
  ) |>
  ggplot(aes(x = x, y = value, color = lambda)) +
    geom_col() +
    facet_wrap(~lambda, ncol = 1) +
    theme(legend.position = "none")
```

### Generating random values with `rpois()`

```{r}
#| title: random normal
#| warning: false

n <- 10000
tibble(
  "lambda = 1" = rpois(n, lambda = 1),
  "lambda = 4" = rpois(n, lambda = 4),
  "lambda = 8" = rpois(n, lambda = 8)
)  |>
  pivot_longer(
    cols = starts_with("lambda = "),
    names_to = "lambda",
    values_to = "value"
  ) |>
  ggplot(aes(value, color = lambda)) + 
    geom_bar() +
    xlim(0, 15) +
    facet_wrap(~lambda, ncol = 1) +
    theme(legend.position = "none")
```

[web-catalina]: https://catalina.lpl.arizona.edu/faq
[concept-randomness]: https://exercism.org/tracks/r/concepts/randomness