---
title: "Statistical Distributions in R"
format: 
  html:
    code-fold: true
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    geometry:
      - top=15mm
      - bottom=15mm
      - heightrounded
---

We will need the Tidyverse libraries and some default settings:

```{r}
#| title: imports
#| warning: false

library(tidyverse)

set_theme(theme_gray(base_size = 14))
```

## Normal/Gaussian Distribution

### The Normal Probability Density Function

```{r}
#| title: normal PDF

tibble(
  "x" = seq(-4, 4, length.out = 1000),
  "SD = 0.5" = dnorm(x, sd = 0.5),
  "SD = 1" = dnorm(x, sd = 1),
  "SD = 2" = dnorm(x, sd = 2)
)  |>
  pivot_longer(
    cols = starts_with("SD = "),
    names_to = "SD",
    values_to = "value"
  ) |>
  ggplot(aes(x = x, y = value, color = SD)) +
    geom_line() +
    facet_wrap(~SD, ncol = 1) +
    theme(legend.position = "none")
```

### The Normal Cumulative Distribution Function

```{r}
#| title: normal CDF

tibble(
  "x" = seq(-4, 4, length.out = 1000),
  "SD = 0.5" = pnorm(x, sd = 0.5),
  "SD = 1" = pnorm(x, sd = 1),
  "SD = 2" = pnorm(x, sd = 2)
)  |>
  pivot_longer(
    cols = starts_with("SD = "),
    names_to = "SD",
    values_to = "value"
  ) |>
  ggplot(aes(x = x, y = value, color = SD)) +
    geom_line() +
    facet_wrap(~SD, ncol = 1) +
    theme(legend.position = "none")
```

### Generating random values with `rnorm()`

```{r}
#| title: random normal

n <- 10000
tibble(
  "SD = 0.5" = rnorm(n, sd = 0.5),
  "SD = 1" = rnorm(n, sd = 1),
  "SD = 2" = rnorm(n, sd = 2)
)  |>
  pivot_longer(
    cols = starts_with("SD = "),
    names_to = "SD",
    values_to = "value"
  ) |>
  ggplot(aes(value, color = SD)) + 
    geom_freqpoly(binwidth = 0.05) +
    facet_wrap(~SD, ncol = 1) +
    theme(legend.position = "none")
```

## Poisson Distribution

### The Poisson Probability Density Function

```{r}
#| title: Poisson PDF

tibble(
  "x" = 0:15,
  "lambda = 1" = dpois(x, lambda = 1),
  "lambda = 4" = dpois(x, lambda = 4),
  "lambda = 8" = dpois(x, lambda = 8)
)  |>
  pivot_longer(
    cols = starts_with("lambda = "),
    names_to = "lambda",
    values_to = "value"
  ) |>
  ggplot(aes(x = x, y = value, color = lambda)) +
    geom_col() +
    facet_wrap(~lambda, ncol = 1) +
    theme(legend.position = "none")
```

### The Poisson Cumulative Distribution Function

```{r}
#| title: Poisson CDF

tibble(
  "x" = 0:15,
  "lambda = 1" = ppois(x, lambda = 1),
  "lambda = 4" = ppois(x, lambda = 4),
  "lambda = 8" = ppois(x, lambda = 8)
)  |>
  pivot_longer(
    cols = starts_with("lambda = "),
    names_to = "lambda",
    values_to = "value"
  ) |>
  ggplot(aes(x = x, y = value, color = lambda)) +
    geom_col() +
    facet_wrap(~lambda, ncol = 1) +
    theme(legend.position = "none")
```

### Generating random values with `rpois()`

```{r}
#| title: random normal

n <- 10000
tibble(
  "lambda = 1" = rpois(n, lambda = 1),
  "lambda = 4" = rpois(n, lambda = 4),
  "lambda = 8" = rpois(n, lambda = 8)
)  |>
  pivot_longer(
    cols = starts_with("lambda = "),
    names_to = "lambda",
    values_to = "value"
  ) |>
  ggplot(aes(value, color = lambda)) + 
    geom_bar() +
    facet_wrap(~lambda, ncol = 1) +
    theme(legend.position = "none")
```

